Network Cascades

### 2016-09-22 paper/paper.tex
Incorporate review 1's suggestions

One way to explain these phenomena is \emph{network cascades}. To stick with the high-level example above, assume that one person starts to use some communication technology. Then they tell their friends about it, convincing some of them to also begin using the application to communicate. Those friends to the same with their own friends, and so on and so forth. Given a suitably connected social network and a high enough ``convincability'' of its members, the usage will cascade through the entire network and eventually spread to all, or at least most, of its members. On the other hand, if the network does not have the required connectivity or if important bridge members cannot be convinced of the switch, the cascade may be halted and will not cause a global change in adoption.

To give another example, the model of network cascades can help to explain how some cultural artifacts like books or movies can emerge out of obscurity, while other artifacts never receive the attention of the general public, although being qualitatively comparable on an objective level. The basic assumption here is that the audience will make the decision of whether to give an artifact (a product) attention based on their observation of their social neighbors, rather than on their personal information. ``If all other people in my group of friends are reading a book then I too should read that book''.

@@

In our example, we choose $z$ to be 2. This means that the graph of the degree distribution for $p_k$ has its peak at 2 and that we would expect that most nodes have 2 neighbors. Plugging that into the distribution function tells us that the probability for this case (2 neighbors) is about $27\%$, and the probability for 1 neighbor is $27\%$, $18\%$ for 3 neighbors, $9\%$ for 4 and $3\%$ for 5, and so on. According to this, a ``realistic'' network for our potential WhatsApp users might have an connection between Alice and Bob, between Bob and Charley, and Charley and Dave, leaving Alice and Dave with each one neighbour, and Bob and Charley with each two neighbors. Figure \ref{fig:abcd} shows the graph for this setup.

@@

In the example, there are $n = 4$ nodes. With $z = 2$ $p$ is $\frac{2}{4} = 0.5$, so we would expect half of all possible edges to exist. There are 6 possible edges in the network (this is excluding edges from a node to itself), and we chose to create 3, so our example indeed conforms perfectly to the Erdős-Rényi-Model.

@@

Now the model is simulated over a time $t$, starting with $0$. At a random point in time, the state of a single node (the seed) is changed to $1$. This is analogous to the first person starting to use the application.

@@

In the example with Alice and the others, we apply the first probability distribution from above. Let us say that Alice and Bob are assigned a threshold of $\phi_{Alice} = \phi_{Bob} = 0.5$ (for which the probability is $20\%$), Charley is assigned $\phi_{Charley} = 0.3$ and Dave $\phi_{Dave} = 0.9$. This means that Dave is very hard to ``convince'', at least $90\%$ of his neighbors must use WhatsApp before he decides to use it (switch his state to $1$), whereas Charley already switches if only $30\%$ of his neighbors adopt the technology. Since Dave has only 1 neighbor, Charley, he switches as soon as Charley switches (in this case, $100\%$ of his neighbors would have switched). Dave has two neighbors, Bob and Dave, but only needs $30\%$ to switch, so it suffices if either Bob or Dave switch (that would be $50\%$ of his neighbors).

@@

What exactly constitutes as a successful global cascade is defined by $\frac{\textmd{Number of agents}}{\textmd{Agents in state } 1} \geq T$, where $T \in [0, 1]$ is some arbitrarily chosen threshold value. For example, $T = 1$ would require every last agent to have changed its state to $1$, which is difficult to achieve in a large random graph. A value like $T = 0.9$ would only require a 90\% majority, but still intuitively result in a cascade that was ``sufficiently global''.

@@

In this configuration, nothing actually occurs yet however, since all states are $0$. All agents do periodically check their neighbors and calculate if they should change their own state, but of course this always results in no change. To actually begin the simulation proper, we must create the initial impulse. That is, we change the state of one node to $1$ at a random point in time. In this case, we have chosen the node with threshold $\phi = 1$. The new state can be seen in figure \ref{fig:model2}.

@@

As should be obvious by now, the final two nodes will also change their state at this point. First the node with $\phi = 0.5$ and finally the node with $\phi = 0.8$. At that point, the entire network is in a stable, unchanging state, as can be seen in figure \ref{fig:model5}. Since the entire network has changed its state to a uniform $1$, the cascade successfully worked its way from a single impulse to covering the entire network.

@@

Watts claims that global cascades are only possible if \textquote{the largest vulnerable cluster percolates}. This is called the \emph{cascade condition}.

@@

A node is vulnerable if its threshold is low enough so that only one of its neighbors needs to switch state for it to switch state itself. Such nodes can be thought of as ``early adopters'' of a technology, with the initial seed being an ``innovator''. A vulnerable cluster is a connected set of nodes that are ``early adopters'' (that are vulnerable) adjacent to an ``innovator''. This cluster must span a sufficiently (Watts never explicitly states what ``sufficient'' in this case means) large portion of the network (percolate) for the cascade condition to be true.

@@

To illustrate this, let us choose those distributions and see what happens. We will construct a uniform graph with a homogeneous threshold. Uniform graphs have a Poisson degree distribution $p_k = \frac{e^{-z}z^k}{k!}$ with $z$ being the mean (the peak) of the distribution. The fixed threshold is easily modeled: The probability for a node being vulnerable is $\rho_k = P(\phi \leq \frac{1}{k})$, but since $\phi$ (the threshold) in this case is fixed, we just need to use $1$ for $\phi \leq \frac{1}{k}$ and $0$ for $\phi > \frac{1}{k}$ (and $1$ for $k = 0$).

@@

The dotted line in the diagram is the result of empirical testings where random networks where generated for the same distributions and the two parameters and tested, i.e. simulated. It is obvious that the calculated and the empirical phase transitions align (with reasonable precision), which validates the cascade condition inequation.

@@

In the above section \ref{fig:cross-section} was introduced as a means to compare the properties of the cascade window for a given threshold. Watts' interpretation of the results is as follows:

@@

An introduction to generating functions can be found in \cite{wilf1994generatingfunctionology} (this version of the book is referenced by Watts et. al. in his paper below).

>>>>

One way to explain these phenomena is \emph{network cascades}. To stick with the high-level example above, assume that one person starts to use some communication technology. Then they tell their friends about it, convincing some of them to also begin using the application to communicate. Those friends do the same with their own friends, and so on and so forth. Given a suitably connected social network and a high enough ``convincability'' of its members, the usage will cascade through the entire network and eventually spread to all, or at least most, of its members. On the other hand, if the network does not have the required connectivity or if important bridge members cannot be convinced of the switch, the cascade may be halted and will not cause a global change in adoption.

To give another example, the model of network cascades can help to explain how some cultural artifacts like books or movies can emerge out of obscurity, while other artifacts never receive the attention of the general public, although being qualitatively comparable on an objective level. The basic assumption here is that the audience will make the decision of whether to give an product or ``artifact'' attention based on their observation of their social neighbors, rather than on their personal information. ``If all other people in my group of friends are reading a book then I too should read that book''.

@@

In our example, we choose $z$ to be 2. This means that the graph of the degree distribution for $p_k$ has its peak at 2 and that we would expect that most nodes have 2 neighbors. Plugging that into the distribution function tells us that the probability for this case of 2 neighbors is about $27\%$, and the probability for 1 neighbor is $27\%$, $18\%$ for 3 neighbors, $9\%$ for 4 and $3\%$ for 5, and so on. According to this, a ``realistic'' network for our potential WhatsApp users might have an connection between Alice and Bob, between Bob and Charley, and Charley and Dave, leaving Alice and Dave with each one neighbour, and Bob and Charley with each two neighbors. Figure \ref{fig:abcd} shows the graph for this setup.

@@

In the example, there are $n = 4$ nodes. With $z = 2$ $p$ is $\frac{2}{4} = 0.5$, so we would expect half of all possible edges to exist. There are 6 possible edges in the network, excluding edges from a node to itself, and we chose to create 3, so our example indeed conforms perfectly to the Erdős-Rényi-Model.

@@

Now the model is simulated over a time $t$, starting with $0$. At a random point in time, the state of a single node, called ``the seed'', is changed to $1$. This is analogous to the first person starting to use the application.

@@

In the example with Alice and the others, we apply the first probability distribution from above. Let us say that Alice and Bob are assigned a threshold of $\phi_{Alice} = \phi_{Bob} = 0.5$ (for which the probability is $20\%$), Charley is assigned $\phi_{Charley} = 0.3$ and Dave $\phi_{Dave} = 0.9$. This means that Dave is very hard to ``convince'', at least $90\%$ of his neighbors must use WhatsApp before he decides to use it i.e. switch his state to $1$, whereas Charley already switches if only $30\%$ of his neighbors adopt the technology. Since Dave has only 1 neighbor, Charley, he switches as soon as Charley switches: In this case, $100\%$ of his neighbors would have switched. Dave has two neighbors, Bob and Dave, but only needs $30\%$ to switch, so it suffices if either Bob or Dave switch. This would be $50\%$ of his neighbors.

@@

What exactly constitutes as a successful global cascade is defined by $NumberOfAgents/AgentsInState1 \geq T$, where $T \in [0, 1]$ is some arbitrarily chosen threshold value. For example, $T = 1$ would require every last agent to have changed its state to $1$, which is difficult to achieve in a large random graph. A value like $T = 0.9$ would only require a 90\% majority, but still intuitively result in a cascade that was ``sufficiently global''.

@@

In this configuration, nothing actually occurs yet however, since all states are $0$. All agents do periodically check their neighbors and calculate if they should change their own state, but of course this always results in no change. To actually begin the simulation properly, we must create the initial impulse. That is, we change the state of one node to $1$ at a random point in time. In this case, we have chosen the node with threshold $\phi = 1$. The new state can be seen in figure \ref{fig:model2}.

@@

As should be obvious by now, the final two nodes will now also change their state. First the node with $\phi = 0.5$ and finally the node with $\phi = 0.8$. At that point, the entire network is in a stable, unchanging state, as can be seen in figure \ref{fig:model5}. Since the entire network has changed its state to a uniform $1$, the cascade successfully worked its way from a single impulse to covering the entire network.

@@

Watts claims that global cascades are only possible if \textquote{the largest vulnerable cluster percolates}\cite{simplemodel}. This is called the \emph{cascade condition}.

@@

A node is vulnerable if its threshold is low enough so that only one of its neighbors needs to switch state for it to switch state itself. Such nodes can be thought of as ``early adopters'' of a technology, with the initial seed being an ``innovator''. A vulnerable cluster is a connected set of nodes that are ``early adopters'' (that are vulnerable) adjacent to an ``innovator''. This cluster must span a sufficiently large portion of the network for the cascade condition to be true. ``Spanning the network'' is called ``percolating'' int the paper. Watts never explicitly states what ``sufficient spanning'' in this case means, unfortunately.

@@

To illustrate this, let us choose those distributions and see what happens. We will construct a uniform graph with a homogeneous threshold. Uniform graphs have a Poisson degree distribution $p_k = \frac{e^{-z}z^k}{k!}$ with $z$ being the mean of the distribution. The fixed threshold is easily modeled: The probability for a node being vulnerable is $\rho_k = P(\phi \leq \frac{1}{k})$, but since the threshold $\phi$ in this case is fixed, we just need to use $1$ for $\phi \leq \frac{1}{k}$ and $0$ for $\phi > \frac{1}{k}$ (and $1$ for $k = 0$).

@@

The dotted line in the diagram is the result of empirical testings where random networks where generated for the same distributions and the two parameters and tested, i.e. simulated. It is obvious that the calculated and the empirical phase transitions align with reasonable precision, which validates the cascade condition inequation.

@@

In the above figure \ref{fig:cross-section} was introduced as a means to compare the properties of the cascade window for a given threshold. Watts' interpretation of the results is as follows:

@@

An introduction to generating functions can be found in \cite{wilf1994generatingfunctionology}. Tthis version of the book is referenced by Watts et. al. in his paper below.
